---
title: "NCAA Basketball Analytics: What Drives Team Success?"
author: "Zach Quicksall"
date: today
format:
  html:
    math: mathjax
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
source("R/packages.R")
source("R/functions.R")
set.seed(2026)

tar_load(team_summary)
tar_load(season_summary)
tar_load(kp_four_factors)

combined_games <- team_summary %>%
  left_join(
    kp_four_factors,
    by = c("season", "season_type", "game_id", "team_id", "team_name")
  )
```

College basketball performance is often summarized by win-loss record, but deeper insights come from possession-based analytics.

In this analysis, we decompose team performance into efficiency metrics and the Four Factors framework introduced by Dean Oliver.

We aim to answer:

- What differentiates elite offensive teams?
- Which factors are most associated with success?
- Do teams succeed in different ways?

## What is the Overall Picture?

Teams in the top-right quadrant combine elite offense and defense, while others specialize on one side of the ball.

```{r}
annot <- season_summary %>%
  filter(offensive_rating > 500, defensive_rating > 500)

season_summary %>%
  ggplot(aes(x = offensive_rating, y = defensive_rating)) +
  geom_point(alpha = 0.6) +
  ggrepel::geom_label_repel(
    aes(
      x = offensive_rating,
      y = defensive_rating,
      label = team_name),
    data = annot) +
  geom_hline(yintercept = mean(season_summary$defensive_rating, na.rm = TRUE), linetype = "dashed") +
  geom_vline(xintercept = mean(season_summary$offensive_rating, na.rm = TRUE), linetype = "dashed") +
  labs(
    title = "Offensive vs Defensive Efficiency",
    x = "Offensive Rating",
    y = "Defensive Rating"
  ) +
    theme_minimal()
```

## Not All Possessions Are Equal

```{r}
season_summary %>%
  ggplot(aes(x = total_possessions, y = total_points_scored / games)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "Possessions vs Points per Game",
    x = "Possessions",
    y = "Points per Game"
  )
```

## Efficiency Separates Teams

```{r}
combined_games %>%
  ggplot(aes(x = tov_pct, y = offensive_rating)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs() +
  theme_minimal()
```

## Turnovers Are a Clear Limiter

```{r}
combined_games %>%
  ggplot(aes(x = tov_pct, y = offensive_rating)) +
  geom_point(alpha = 0.4) +
  labs(x = "Turnover Percentage Per Possession (%)", y = "Offensive Rating") +
  theme_minimal()
```

## Offensive Rebounding Creates Second Chances

```{r}
combined_games %>%
  ggplot(aes(x = orb_rate, y = offensive_rating)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Offensive Rebound Rate (%)", y = "Offensive Rating") +
  theme_minimal()
```

## Free Throws Reflect Play Style

```{r}
combined_games %>%
  ggplot(aes(x = ft_rate, y = offensive_rating)) +
  geom_point(alpha = 0.4) +
  labs(x = "Free Throw Rate (%)", y = "Offensive Rating") +
  theme_minimal()
```

## Multiple Paths to Success

```{r}
#combined_season %>%
#  ggplot(aes(x = tov_pct, y = orb_rate, color = net_rating)) +
#  geom_point() +
#  scale_color_viridis_c()
```

## Efficiency Wins Overall

```{r}

```

## Teams That Break the Pattern

```{r}

```